draftFiguresPaper

### 2025-09-11 draftFiguresPaper.Rmd
finalDatasetFigures lets discuss what gets included in the paper, the color scale, and whether we compare nb of papers and pollution levels

datasetClassification <- read.csv("output/dataClassifToPlot.csv")

datasetCountries <- read.csv("dataNew/allgpt-lit_w_countries.csv")

datasetClassificationCountries <- merge(datasetClassification, datasetCountries %>% 
                                          select(X, Year, country) %>% 
                                          rename(ID = X),
                                        by = "ID") %>% 
  select(-reader)

@@

# 1) split into one row per country
datasetClassificationCountries_long <- datasetClassificationCountries %>%
  separate_rows(country, sep = ";\\s*") %>%
  mutate(countries = str_trim(country))

# fix country names 
library(countrycode)

@@

pCountryPaperIncome <- papers_by_country %>%
  slice_max(n_papers, n = 20) %>%                     # top 30
  mutate(
    # 1) reorder by paper count (so bars go largest→smallest),
    country = fct_reorder(country, n_papers),
    # 2) then shove "Unknown" to the very end
    country = fct_relevel(country, "Unknown", after = 0)
  ) %>%
  ggplot(aes(x = country, y = n_papers, fill = Income.group)) +
    geom_col() +
    coord_flip() +
    scale_fill_discrete(na.value = 'grey')+
    labs(
      x = "Country",
      y = "Number of papers",
      title = "Top 20 Countries by Paper Count"
    ) +
    theme_minimal()

@@

### add  monitor data to compare
# load the monitor dataset and subset it
monitors <- read.csv("dataNew/mergedMonitors_v7.csv")

# keep only reference grade monitors
monitorsRef <- subset(monitors, isMonitor)

# keep only active (different datasets have different ways of dealing with that)
# airnow
# no subset because no precise info on when active

# openaq
monitorsRef$diffDays <- as.numeric(difftime(monitorsRef$lastUpdated, monitorsRef$firstUpdated, units = "days"))

monitorsRef <- subset(monitorsRef, !(diffDays < 30 & firstUpdated < ymd_hms("2025-02-01T00:00:00+00:00") & database == "OpenAQ"))

# waqi
monitorsRef <- subset(monitorsRef, !(occurenceWAQI < 2 & database == "WAQI"))

# WHO I am not sure if we should filter points that are only active for a year
table(monitorsRef$database)

# we probably want to check if the results are consistent if removing embassy monitors
table(monitorsRef$isEmbassy)

monitorsRef_simple <- monitorsRef %>%
  # filter(!isEmbassy) %>% # here to filter embassies
  select(newID, longitude, latitude)

monitors_vect <- vect(monitorsRef_simple, geom=c("longitude", "latitude"), crs=crs(world_no_antarctica))

# 1. intersect: this returns only those points lying in (or on) each polygon,
#    with the polygon’s attributes copied over
pts_in_c <- intersect(monitors_vect, world_no_antarctica)

# 2. now simply tally by the country name
counts2 <- table(pts_in_c$iso_a3)

print(counts2)

# aggregate the intersected points by country NAME
agg <- aggregate(pts_in_c, by = "iso_a3", fun = length) %>% as.data.frame() %>% 
  rename(count = agg_newID) %>% 
  select(iso_a3, count)

# join back to the country layer
world_no_antarctica <- merge(world_no_antarctica, agg, by = "iso_a3", all.x = TRUE)

world_no_antarctica_df <- world_no_antarctica %>% as.data.frame()

world_no_antarctica_df$count <- ifelse(is.na(world_no_antarctica_df$count), 0, world_no_antarctica_df$count)

world_no_antarctica_df %>% ggplot(aes(x = n_papers, y = count))+
  geom_point()+
  geom_smooth(method = "lm")+
  facet_wrap(~income_grp)

# world_no_antarctica_df %>% ggplot(aes(x = n_papers, y = count))+
#   geom_point()+
#   geom_smooth(method = "lm")+
#   facet_wrap(~Income.group)

world_no_antarctica_df %>% ggplot(aes(x = n_papers, y = count))+
  geom_point()+
  geom_smooth(method = "lm")

subset(world_no_antarctica_df, count < 10) %>% ggplot(aes(x = n_papers, y = count))+
  geom_point()+
  geom_smooth(method = "lm")

#### vs mean levels of pollution !!! to be done ####

@@

>>>>

datasetClassification <- read.csv("output/dataClassifToPloto4mini.csv")

datasetCountries <- read.csv("output/fullClassifCountryo4mini.csv")

datasetCountries <- datasetCountries %>% 
  select(ID, countryPartialF) %>% 
  rename(country = countryPartialF)

datasetClassificationCountries <- merge(datasetClassification, datasetCountries,
                                        by = "ID") %>% 
  select(-reader)

@@

# 1) split into one row per country
datasetClassificationCountries_long <- datasetClassificationCountries %>%
  separate_rows(country, sep = ",\\s*") %>%
  mutate(countries = str_trim(country))

datasetClassificationCountries_long$countries <- ifelse(datasetClassificationCountries_long$countries == "Turkyie", "Turkey",
                                                        ifelse(datasetClassificationCountries_long$countries %in% c("Wales", "England"), "United Kingdom",
                                                               datasetClassificationCountries_long$countries))
# fix country names 
library(countrycode)

@@

pCountryPaperIncome <- papers_by_country %>%
  slice_max(n_papers, n = 20) %>%                     # top 30
  mutate(
    # 1) reorder by paper count (so bars go largest→smallest),
    country = fct_reorder(country, n_papers)
  ) %>%
  ggplot(aes(x = country, y = n_papers, fill = Income.group)) +
    geom_col() +
    coord_flip() +
    scale_fill_discrete(na.value = 'grey')+
    labs(
      x = "Country",
      y = "Number of papers",
      title = "Top 20 Countries by Paper Count"
    ) +
    theme_minimal()

@@

@@

# data with ID year 
datasetYear <- read.csv("dataNew/datasetToClassifyGPT.csv") %>% 
  select(ID, Year)

datasetClassificationCountries_long_clean <- merge(datasetClassificationCountries_long_clean,
                                                   datasetYear,
                                                   by = "ID", all.x = T, all.y = F)
